---
title: "Gov 50 Final Project"
author: "Eric Thompson"
description: "Military Service Impact on Voting Habits when Military Veterans are Running for Federal Office"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

  For this research paper, the main question is: Are military-affiliated voters more likely to vote for fellow military veterans running for federal office (House of Representatives and Senate) than voters that have no military affiliation? In this study, I seek to show that voters with prior or current military affiliation are more likely to vote for candidates that are themselves military veterans than voters that have no military affiliation through their own service. My hypotheses are as follows:
  
  Null: there is no difference in voting patterns for military veteran candidates between voters that have military service and voters that do not.
  
  Alternative: military-affiliated voters are more likely to vote for fellow veterans than non-military-affiliated voters.
  
  This research question and hypothesis are important because they could have impacts on political races going into the future. As the wars in the Middle East that have been being fought for the last 20-some years come to a close, more military veterans will be getting out of the military and going into other lines of work. One profession that draws a fair amount of military talent is politics. It would be a huge leg up if military veteran candidates were more likely to get a portion of the vote just based off their prior work experience, regardless of political party affiliation, where their opponent does not have that luxury. 
  
## Data Section
  
  My data comes from a CCES 2022 survey, which includes a person's history with the military (i.e. currently serving, used to serve, never served), their political leanings, the candidates on which they voted for each election (House, Senate, President, Governor), and who the other candidate(s) were in those elections. 
  
  The explanatory variables of interest would be 1) whether a candidate is a veteran or not and 2) whether or not a voter has any military affiliation or not, with military affiliation defined as either having served in the military in the past or currently serving. In order to determine which candidates are veterans, I will have to bring in a separate data set that contains all of the veterans that ran for House representative and Senate positions in 2022 and from there create a binary variable that indicates if a candidate is a veteran, then use that binary variable to create another one that determines whether a voter voted for the veteran candidate or not. Currently the military affiliation is split up into 5 categories, but it will be changed to only be a 1 or 0 value, indicating if you have ever served in the military or not. 
  
  The outcome variable would be the percentage of the time that military-affiliated voters voted for veteran candidates compared to non-military-affiliated voters. If we see that military-affiliated voters vote more often for veteran candidates than non-veterans and it is statistically significant, this would be supportive of my hypothesis. A pattern that would disprove my hypothesis would be if we see that military-affiliated voters were no more likely to vote for military veteran candidates that were running for office than voters who have no military affiliation were. 
  Since this study is going to compare the voting trends between two separate groups, where the data was collected from both groups at one point in time, the research design is a cross-sectional design, looking specifically at a difference-in-means between the two groups.

```{r}
library(tidyverse)
library(dplyr)
library(broom)
library(infer)
library(modelsummary)
```


```{r}
# load cces data
cces <- read_csv("CCES22_edited.csv")
vet_cand_names <- read_csv("Vet_Cand_Names.csv")
```

```{r}
cces <- cces |> filter(CC22_401 == 5) |>
  mutate(military_mem = if_else(milstat_1 == 1 | milstat_3 == 1, "Military", "Non-Military"))

# create visual
mil_member_count_bar <- cces |> group_by(military_mem) |>
  summarize(n = n()) |>
  ggplot(mapping = aes(x = military_mem, y = n, fill = military_mem)) +
  geom_col() +
  labs(
    x = "Military Affiliation of Voters",
    y = "Count of Voters",
    title = "Number of Military Affiliated and Non-Military Affiliated Voters in Sample",
    source = "CCES 2022 Midterm Election Data, Military Times"
  ) + scale_fill_discrete(name = "Voter Military Affiliation") +
  geom_text(aes(label = n),size = 4,
    vjust = 1.5, position = position_dodge(.9)) +
  scale_fill_manual(values = c("Non-Military" = "steelblue1", "Military" = "indianred1"))

mil_member_count_bar
```

Here we can see that military-affiliated voters are largely outnumbered by voters that have no military affiliation, which is to be expected based off of demographic trends in the United States where the data was pulled from. The number of people in the US that have never served in the military is far greater than the number of people that have, so these numbers appear to be representative of the country as a whole. 

```{r}
cces <- cces |>
  mutate(vet_house_cand = if_else(HouseCand1Name_post %in% vet_cand_names$Name | HouseCand2Name_post %in% vet_cand_names$Name, "Veteran in Race", "No Veteran in Race")) |>
  mutate(vet_sen_cand = if_else(SenCand1Name_post %in% vet_cand_names$Name | SenCand2Name_post %in% vet_cand_names$Name, "Veteran in Race", "No Veteran in Race")) |>
  mutate(vet_on_ballot = if_else(vet_sen_cand == "Veteran in Race" | vet_house_cand == "Veteran in Race", "Veteran on Ballot", "No Veterans on Ballot"))

cces |> group_by(vet_on_ballot) |>
  summarize(n = n(), na.rm = TRUE) |>
            mutate(prop = n / sum(n)) |>
   ggplot(mapping = aes(x = vet_on_ballot, y = prop, fill = vet_on_ballot)) +
   geom_col() +
  labs(
    x = "Were Veterans on the Voting Ballot",
    y = "Proportion of Total Ballots",
    title = "Proportion of respondents that had at least one veteran on their voting ballot",
  ) +
  scale_fill_discrete(name = "Ballot Indicator") +
  geom_text(aes(label = prop),size = 4,
    vjust = 1.5, position = position_dodge(.9)) +
  scale_fill_manual(values = c("No Veterans on Ballot" = "steelblue1", "Veteran on Ballot" = "indianred1"))
```

This visual shows the proportion of observations in the data set where the voter voted in at least one election where a veteran candidate was available to vote for on their ballot. While the proportion indicates a little less than half of the observations had this option, it still leaves a large number of observations (18276) with which to conduct the analysis of interest to this study. 

## Results Section

```{r}
cces <- cces |>
  mutate(vet_house_cand = if_else(HouseCand1Name_post %in% vet_cand_names$Name | HouseCand2Name_post %in% vet_cand_names$Name, "Veteran in Race", "No Veteran in Race")) |>
  mutate(vet_sen_cand = if_else(SenCand1Name_post %in% vet_cand_names$Name | SenCand2Name_post %in% vet_cand_names$Name, "Veteran in Race", "No Veteran in Race")) |>
  mutate(vet_on_ballot = if_else(vet_sen_cand == "Veteran in Race" | vet_house_cand == "Veteran in Race", "Veteran on Ballot", "No Veterans on Ballot"))

```

```{r}
# create indicator for if a person voted for a military veteran candidate
cces <- cces |>
  mutate(voted_for_vet = if_else((HouseCand1Name_post %in% vet_cand_names$Name & CC22_412 == 1) | (HouseCand2Name_post %in% vet_cand_names$Name & CC22_412 == 2) | (SenCand1Name_post %in% vet_cand_names$Name & CC22_411 == 1) | (SenCand2Name_post %in% vet_cand_names$Name & CC22_411 == 2), 1, 0))

# subset to just respondents that had a veteran on the ballot
cces_vet_cands <- cces |> filter(vet_on_ballot == "Veteran on Ballot")

vet_cand_tab <- cces_vet_cands |>
  group_by(military_mem, voted_for_vet) |>
  drop_na(voted_for_vet) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n)) |>
  select(-n) |>
  pivot_wider(
    names_from = voted_for_vet,
    values_from = prop
  )

vet_cand_tab <- knitr::kable(
  vet_cand_tab,
  col.names = c("Voter Military Association", "Share that Voted For Non-Veteran Candidate", "Share that Voted for Veteran Candidate")
) 

vet_cand_tab

# calculate the average treatment effect
ATE <- cces_vet_cands |>
  group_by(military_mem) |>
  drop_na(voted_for_vet) |>
  summarize(prop = mean(voted_for_vet)) |>
  pivot_wider(
    names_from = military_mem,
    values_from = prop
  ) |>
  mutate("Average Treatment Effect" = Military - `Non-Military`)

ATE <- knitr::kable(ATE,
             col.names = c("Military-Affiliated Voters", "Non-Military Affiliated Voters", "Average Treatment Effect"))

ATE
```

```{r} 
# create a visual that shows the proportion of military members and veterans that vote for candidates that are veterans compared to non-veteran voters
vet_cand_vote <- cces_vet_cands |>
  mutate(voted_for_vet = if_else(voted_for_vet == 1,
                             "Yes", "No")) |>
  group_by(military_mem, voted_for_vet) |>
  drop_na(voted_for_vet) |>
  summarize(n = n()) |>
  mutate(prop = n / sum(n)) |>
  select(-n) |>
  filter(voted_for_vet == "Yes") |>
  ggplot(mapping = aes(x = military_mem, y = prop, fill = military_mem)) +
  geom_col(position = "dodge")  +
  labs(
    x = "Military Affiliation",
    y = "Proportion of Group",
    title = "Proportion of Group that Voted for the Military Veteran Candidate",
    source = "CCES 2022 Midterm Election Data, Military Times"
  ) + scale_fill_discrete(name = "Voter Military Affiliation") +
  geom_text(aes(label = prop),size = 4,
    vjust = 1.5, position = position_dodge(.9)) +
  scale_fill_manual(values = c("Non-Military" = "steelblue1", "Military" = "indianred1"))

#display visual
vet_cand_vote
```

```{r}
# reorder variables so we can see the military member regression coefficient in the output
cces_vet_cands$military_mem <- factor(cces_vet_cands$military_mem, levels = c("Non-Military", "Military"))

# regress military-affiliation on voting for a military veteran candidate
reg_vote <- lm(voted_for_vet ~ military_mem, data = cces_vet_cands)

# creating variable labels for regression output
var_lables <- c(
  "(Intercept)" = "Intercept",
  "military_memMilitary" = "Military-Affiliated Voter"
)

# creating a more visually appealing regression output
regression_model <- modelsummary(
  reg_vote,
  statistic = c("s.e. = {std.error}",
                "p = {p.value}"),
  coef_map = var_lables,
  fmt = 4, 
  title = "Regression Output of Military-Affiliation on Voting for Veteran Candidates", 
  gof_map = c("nobs"))

regression_model
```


```{r}
# count how many military and non-military voters are in the data set
voter_counts <- cces_vet_cands |> 
  count(military_mem)

cces_vet_cands_test <- cces_vet_cands |>
  mutate(vote = if_else(voted_for_vet == 1, "Voted", "Didn't Vote")) |>
  select(vote, voted_for_vet, military_mem) |>
  na.omit()

# estimate the ATE
est_ate <- cces_vet_cands_test |>
  specify(vote ~ military_mem, success = "Voted") |>
  calculate(stat = "diff in props", order = c("Military", "Non-Military"))

# save the standard error to add to table later
se_diff <- 0.010918

# calculate z-score
z_score = est_ate / se_diff

# save p-value to add to table later
p_val <- 0.0583

# calculate 95% confidence interval
ci_upper <- est_ate + (se_diff * 1.96)
ci_lower <- est_ate - (se_diff * 1.96)

# create a data frame of important stats
values <- data.frame(p_val, est_ate, se_diff, z_score, ci_lower$stat, ci_upper$stat)

# make a more visually appealing output
value_table <- knitr::kable(values,
             col.names = c("P-Value", "Average Treatment Effect", "Standard Error", "Z-Score", "Lower 95% Confidence Interval", "Upper 95% Confidence Interval"))
value_table
```

  These tables and plots above show the difference-in-means between voters that have either served in the military or currently do (Military-Affiliated Voters) vs. voters that have never served in the military (Non-Military Affiliated Voters). The estimated average treatment effect suggests that military-affiliated voters are 2.07 percentage points more likely to vote for candidates that are military veterans themselves than non-military affiliated voters are. 
  
  While the coefficient of interest tells us that the treatment of being in the military makes you 2.07 percentage points more likely to vote for candidates that are themselves military veterans than voters that have no military affiliation, we should not interpret this treatment as causal. Using an alpha of 0.05 and looking at the p-value of the regression, as well as the p-values coming from the permutation test and the calculations using the standard error of the difference-in-means, all of them are above 0.05. While all of the p-values hovered just above the significance cutoff of 0.05, they were still over that threshold, meaning that we fail to reject the null hypothesis that there is no difference in voting rates for veteran candidates between military-affiliated voters and voters with no military affiliation. 
  
  Additionally, the 95% confidence interval of (-0.0007, 0.04207) is another reason why the results should not be interpreted causally. Since 0 is within the confidence interval, even though it is at the extremely low end, it cannot be ruled out within our chosen confidence parameters that there is no effect on voting for military veterans from having served in the military.

```{r}
set.seed(02138)

# Permutation Test
null_dist <- cces_vet_cands_test |>
  specify(vote ~ military_mem, success = "Voted") |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "diff in props", order = c("Military", "Non-Military"))

# calculate p-value for permutation test
ate_pval <- null_dist |>
  get_p_value(obs_stat = est_ate, direction = "both")

# visual output of the permutation test sample distribution
null_dist |>
  visualize() +
  shade_p_value(obs_stat = est_ate, direction = "both") +
  labs(
    x = "Sample Average Diff-in-Means",
    y = "Count of Samples",
    title = "Permutation Test Sample Distribution"
  )

# calculate 95% confidence interval of the sampling distribution
ci <- null_dist |>
  get_ci(level = 0.95)
knitr::kable(ci)

# display p-value
knitr::kable(ate_pval)
```

Additionally, to increase the confidence that we cannot reject the null hypothesis, even though the results are close, the p-value from the permutation test is still above our alpha of 0.05. Therefore, based off of our chosen alpha and the results of the permutation test, we cannot rule out that the 0.0207 estimated average treatment effect that we observed in our data set happened by chance, and that there really is no difference between military-affiliated voters and voters with no military affiliation when it comes to voting for candidates that are themselves military veterans. 


## Conclusion

In conclusion, from our sample of CCES data from the 2022 midterm elections, we have estimated that, on average, voters who are currently in the military or have served in the military in the past (military-affiliated voters) are 2.07 percentage points more likely to vote for candidates running for federal office (in this case, House of Representatives and Senate) who are also military veterans than voters that have never served in the military. Using an alpha of 0.05, we fail to reject the null hypothesis that in reality there are no difference in population means between the two voter groups, and that the estimated average treatment effect happened by chance. While the p-values that we extracted from our analyses were close to the threshold for statistical significance, they were not low enough, which means we cannot reject the null. Since this is the case, we cannot confirm that the treatment effect reported above is the causal effect of serving in the military on voting habits. In addition, it is important to consider limitations on this study as well. One major limitation is that arguably the most important aspect of voting habits was not included in this study, and that is party of either the voter or the candidate. This confounder could have large impacts on how voters choose to vote, and could very likely have more of an impact on how they vote than veteran status of a voter and / or candidate. Another aspect that could have skewed these results is that the data is from a midterm election. While midterm elections still have decent voter turnout, it is normally voters that are more dedicated to politics that show up for the midterm elections to vote, whereas the general elections on years where the presidential vote is on the ballot tend to attract more of the general population, so these numbers may be influenced by the type of voter that shows up to midterm elections. Given more time and / or money to improve this analysis, I would incorporate party affiliation of the voters and candidates into the analysis, as well as look at data sets from multiple elections, including a general election, instead of just the single midterm election I looked at in this study.


 


